---
title: "Documentation: Federation Functions"
output:
  pdf_document: default
  html_document: default
---

```{r setup, message=FALSE, warning=FALSE}
source("federated_base_func.R")
source("federated_pool_func.R")
```

In this document, we illustrate how we use the proposed federated inference methods that require only one-time sharing of aggregate data, to obtain the federated point estimates and variance. The data sources are generated with the data simulation scheme as below: 

# Data generation 

In our data generating process, $\mathbf{X}_i = (X_{i 1}, X_{i 2}, X_{i 3})^T \in \mathbb{R}^3$ are i.i.d. samples where each $X_{i,j} \sim \text{unif}(-1, 1)$ is a scalar for $j \in \{1, 2, 3\}$. $W_i$ is a binary treatment variable that follows: 

$$\frac{P(W_i = 1 \mid \mathbf{X}_i)}{P(W_i = 0 \mid \mathbf{X}_i)} = \exp(\gamma_c + \gamma_x^T \mathbf{X}_i)$$
where $\gamma_c = 0.1$ and $\gamma_x = [0.2, 0.3, 0.4]$. $Y_i$ is a binary response variable that follows

$$\frac{P(Y_i = 1\mid \mathbf{X}_i, W_i)}{P(Y_i = 0\mid \mathbf{X}_i, W_i)} = \exp(\beta_c + \beta_w W_i + \beta_x^T \mathbf{X}_i)$$
where $\beta_c = -0.2$, $\beta_w = -0.3$, $\beta_x = [0.5, 0.7, -0.6]$. 


We generate a total 20,000 observations and randomly split these observations into $D=2$ equally-sized data sets. For the illustration purpose, we consider **ATE** as the estimand. 

```{r}
# two equally-sized data sets
D = 2
# entire simulated sample size
N = 20000  
estimand = "ATE"

gamma <- c(0.1, 0.2, 0.3, 0.4) 
beta <- c(-0.2, -0.3, 0.5, 0.7, -0.6) 
```

```{r}
expit <- function(x) {
    return(1/(1 + exp(-x)))
}

GenerateData <- function(N, D, seed=123) {
  set.seed(seed)
  subsample_lst <- list()
  for (ix in 1:D) {
    X1 <- runif(N/D, min = -1, max = 1)
    X2 <- runif(N/D, min = -1, max = 1)
    X3 <- runif(N/D, min = -1, max = 1)
    
    treat <- rbinom(N/D, 1, expit(gamma[1]+gamma[2]*X1+gamma[3]*X2+gamma[4]*X3))
    y0 <- rbinom(N/D, 1, expit(beta[1]+beta[2]*0+beta[3]*X1+beta[4]*X2+beta[5]*X3))
    y1 <- rbinom(N/D, 1, expit(beta[1]+beta[2]*1+beta[3]*X1+beta[4]*X2+beta[5]*X3))
    y <- y1*treat + y0*(1-treat)
    intercept = rep(1, N/D)
    dat <- data.frame(intercept, y, y1, y0, treat, X1, X2, X3)
    dat$subsample <- ix
    subsample_lst[[ix]] <- dat
  }
  pooled <- do.call("rbind", subsample_lst)
  true_ate <- mean(pooled$y1) - mean(pooled$y0)
  pooled$y <- pooled$y1*pooled$treat + pooled$y0*(1-pooled$treat)
  
  return(list(true_ate = true_ate, subsample_lst = subsample_lst, pooled = pooled))
}
```

# MLE 

## data set-specific MLE estimation 

```{r}
temp <- GenerateData(N=20000, D=2, seed=123)
subsample_lst <- temp$subsample_lst
```

If we are interested in any data set-specific MLE estimations, we can use the `est_mle` function and speicfy the outcome (treatment) and covariates vectors, as well as the working candidate models. For example, we can obtain a MLE estimation for the **treatment model** (`treat.reg` set to `TRUE`) with robust variance estimator on the first dataset as follows:

```{r}
# specify the first dataset 
idx <- 1; subsample <- subsample_lst[[idx]]

# specify the working covariates 
covariates <- c("X1", "X2", "X3")   
# specify the working treatment model  
treat.reg.formula <- as.formula(
      paste("treat ~ 0 + intercept + ", paste(covariates, collapse = " + ")))

# get the treatment assignment vector 
treat <- subsample$treat  
# get the design matrix 
X <- model.matrix(treat.reg.formula, data = subsample)
# use robust variance estimator 
robust <- TRUE             

# MLE estimation for treatment model 
result <- est_mle(this.y=treat, this.X=X, robust=TRUE, treat.reg=TRUE, treat.reg.formula = treat.reg.formula)
```

This result is a list of MLE estimates including estimated coefficients: 

```{r}
result$est
```

The variance covariance matrix (robust sandwich form as we have specified, `result$V`): 

```{r}
result$V
```

Besides, it returns the gradient (`result$grad`), outer product of the gradient (`result$outer.grad`), hessian (`result$hessian`) of the log likelihood function evaluated at the MLE estimates, and the estimated probabilities (`result$est.prob`) of being treated. 

Similarly, we can obtain a MLE estimation for the **outcome model** (`treat.reg` set to `FALSE`(default)) with robust variance estimator on the first dataset as follows:

```{r}
# specify the first dataset 
idx <- 1; subsample <- subsample_lst[[idx]]

# specify the working covariates 
covariates <- c("X1", "X2", "X3")   
# specify the working treatment model  
reg.formula <- as.formula(
  paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))

# get the outcome vector 
y <- subsample$y  
# get the design matrix 
X <- model.matrix(reg.formula, data = subsample)
# use robust variance estimator 
robust <- TRUE             

# MLE estimation for outcome model 
result <- est_mle(this.y=y, this.X=X, robust=TRUE, treat.reg=FALSE, reg.formula = reg.formula)
```

## federation using MLE 

### stable outcome models / treatment models 

We provide an example of federating outcome models, the same approach applied to treatment models by changing the input list of data set-specific response variables as treatment assignments. In this example, we assume stable outcome models across data sets and we use a robust federated variance estimator (`model_misspec=TRUE`). 

```{r}
# specify the working covariates 
# for now we assume stable models, so same set of covariates across data sets 
covariates <- c("X1", "X2", "X3")   

# initialize inputs for federated MLE function
full_y <- vector(mode = "list", length = D)
full_X <- vector(mode = "list", length = D)
full_reg.formula <- vector(mode = "list", length = D)
model_misspec <- TRUE

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  full_y[[ix]] <- subsample$y
  formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_X[[ix]] <- model.matrix(formula, data = subsample)
  full_reg.formula[[ix]] <- formula
}

# restricted federated MLE estimation for outcome model 
result <- poolMLE(full_y=full_y, full_X=full_X, full_reg.formula=full_reg.formula, model_misspec=TRUE)
```

The results is a summary table of pooled estimates, standard errors, t statistics and corresponding p-values. 

```{r}
result
```

### unstable outcome models / treatment models 

In this example, we assume unstable outcome models across data sets (`X2` and `X3` are set as unstable covariates) and we use a robust federated variance estimator (`model_misspec=TRUE`). 

```{r}
# specify the working covariates 
# for now we assume stable models, so same set of covariates across data sets 
covariates <- c("X1", "X2", "X3")
covariates_unstable <- c("X2", "X3")

# initialize inputs for federated MLE function
full_y <- vector(mode = "list", length = D)
full_X <- vector(mode = "list", length = D)
full_reg.formula <- vector(mode = "list", length = D)
model_misspec <- TRUE

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  full_y[[ix]] <- subsample$y
  formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_X[[ix]] <- model.matrix(formula, data = subsample)
  full_reg.formula[[ix]] <- formula
}

# unrestricted federated MLE estimation for outcome model 
result <- poolMLE(full_y=full_y, full_X=full_X, 
                  unstable=TRUE, num_datasets=D, unstable_covars = covariates_unstable, 
                  full_reg.formula=full_reg.formula, model_misspec=TRUE)
```

The results is a summary table of pooled estimates, standard errors, t statistics and corresponding p-values. 

```{r}
result
```

Note that we can also specify different working models and specify the non-overlapped covariates of the working models as the unstable covariates as the following example is showing: 

```{r}
# specify the working covariates 
# for now we assume stable models, so same set of covariates across data sets 
covariates <- list(c("X1", "X2", "X3"), c("X1", "X3"))
covariates_unstable <- c("X2")

# initialize inputs for federated MLE function
full_y <- vector(mode = "list", length = D)
full_X <- vector(mode = "list", length = D)
full_reg.formula <- vector(mode = "list", length = D)
model_misspec <- TRUE

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  full_y[[ix]] <- subsample$y
  formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates[[ix]], collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_X[[ix]] <- model.matrix(formula, data = subsample)
  full_reg.formula[[ix]] <- formula
}

# unrestricted federated MLE estimation for outcome model 
result <- poolMLE(full_y=full_y, full_X=full_X, 
                  unstable=TRUE, num_datasets=D, unstable_covars = covariates_unstable, 
                  full_reg.formula=full_reg.formula, model_misspec=TRUE)
```

```{r}
result
```

# IPW-MLE 

## data set-specific IPW-MLE estimation 

If we are interested in any data set-specific IPW-MLE estimations, we can use the `est_ht` function and speicfy the outcome, treatment, covariates vectors, as well as the working candidate models. For example, we can obtain a IPW-MLE estimation for ATE (`estimand="ATE"`) with estimated propensity scores on the first dataset as follows:

```{r, warning = FALSE}
# specify the first dataset 
idx <- 1; subsample <- subsample_lst[[idx]]

# specify the working covariates 
covariates <- c("X1", "X2", "X3")   
# specify the working treatment model  
treat.reg.formula <- as.formula(
      paste("treat ~ 0 + intercept + ", paste(covariates, collapse = " + ")))
# specify the working outcome model  
reg.formula <- as.formula(
      paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))

# get the outcome vector 
y <- subsample$y  
# get the treatment assignment vector 
treat <- subsample$treat  
# get the design matrix (with treatment column)
X.tilde <- model.matrix(reg.formula, data = subsample)
# get the design matrix (no treatment column)
X <- model.matrix(treat.reg.formula, data = subsample)

# IPW-MLE estimation for treatment model 
result <- est_ht(this.y=y, this.X.tilde=X.tilde, this.X=X, this.treat=treat, 
                 reg.formula=reg.formula, treat.reg.formula = treat.reg.formula, 
                 estimand="ATE", estimated_propensity=TRUE)
```

This result is a list of IPW-MLE estimates including estimated coefficients for outcome model (`result$est`): 

```{r}
result$est
```

The variance covariance matrix (`result$V`):

```{r}
result$V
```

And the estimated probabilities for outcomes (`result$est.prob`). 

## federation using IPW-MLE 

### stable models 

In this example, we assume stable outcome models and stable treatment models across data sets. We assume the propensity scores are estimated (`estimated_propensity=TRUE`) and the target estimand is ATE (`estimand="ATE"`). 

```{r}
# specify the working covariates 
# for now we assume stable models, so same set of covariates across data sets 
covariates <- c("X1", "X2", "X3")   

# initialize inputs for federated IPW-MLE function
full_y <- vector(mode = "list", length = D)
full_X.tilde <- full_X <- vector(mode = "list", length = D)
full_treat <- vector(mode = "list", length = D)
full_reg.formula <- full_treat.reg.formula <- vector(mode = "list", length = D)

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  
  reg.formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))
  treat.reg.formula <- as.formula(
    paste("treat ~ 0 + intercept + ", paste(covariates, collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_treat[[ix]] <- subsample$treat
  full_X[[ix]] <- model.matrix(treat.reg.formula, data = subsample)
  full_X.tilde[[ix]] <- model.matrix(reg.formula, data = subsample)
  full_reg.formula[[ix]] <- reg.formula
  full_treat.reg.formula[[ix]] <- treat.reg.formula
}

# restricted federated IPW-MLE estimation 
result <- poolHT(full_y=full_y, full_X.tilde=full_X.tilde, full_X=full_X, full_treat=full_treat, 
                 unstable=FALSE, 
                 full_reg.formula=full_reg.formula, full_treat.reg.formula=full_treat.reg.formula, 
                 estimand="ATE", estimated_propensity=TRUE)
```

The results is a summary table of pooled estimates, standard errors, t statistics and corresponding p-values. 

```{r}
result
```

### unstable outcome models / treatment models 

In this example, we assume unstable outcome models and unstable treatment models across data sets. We assume the propensity scores are estimated (`estimated_propensity=TRUE`) and the target estimand is ATE (`estimand="ATE"`). 

```{r}
# specify the working covariates 
# for now we assume stable models, sosame set of covariates across data sets 
covariates <- c("X1", "X2", "X3")   
covariates_unstable <- c("X2", "X3")

# initialize inputs for federated IPW-MLE function
full_y <- vector(mode = "list", length = D)
full_X.tilde <- full_X <- vector(mode = "list", length = D)
full_treat <- vector(mode = "list", length = D)
full_reg.formula <- full_treat.reg.formula <- vector(mode = "list", length = D)

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  
  reg.formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))
  treat.reg.formula <- as.formula(
    paste("treat ~ 0 + intercept + ", paste(covariates, collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_treat[[ix]] <- subsample$treat
  full_X[[ix]] <- model.matrix(treat.reg.formula, data = subsample)
  full_X.tilde[[ix]] <- model.matrix(reg.formula, data = subsample)
  full_reg.formula[[ix]] <- reg.formula
  full_treat.reg.formula[[ix]] <- treat.reg.formula
}

# restricted federated IPW-MLE estimation 
result <- poolHT(full_y=full_y, full_X.tilde=full_X.tilde, full_X=full_X, full_treat=full_treat, 
                 unstable=TRUE, unstable_covars = covariates_unstable, num_datasets = D, 
                 full_reg.formula=full_reg.formula, full_treat.reg.formula=full_treat.reg.formula, 
                 estimand="ATE", estimated_propensity=TRUE)
```

The results is a summary table of pooled estimates, standard errors, t statistics and corresponding p-values. 

```{r}
result
```

# AIPW

## data set-specific AIPW estimation 

If we are interested in any data set-specific AIPW estimations, we can use the `est_aipw` function and speicfy the outcome, treatment, covariates vectors, as well as the working candidate models. For example, we can obtain a AIPW estimation for ATE (`estimand="ATE"`) on the first dataset as follows:

```{r}
# specify the first dataset 
idx <- 1; subsample <- subsample_lst[[idx]]

# specify the working covariates 
covariates <- c("X1", "X2", "X3")   
# specify the working treatment model  
treat.reg.formula <- as.formula(
      paste("treat ~ 0 + intercept + ", paste(covariates, collapse = " + ")))
# specify the working outcome model  
reg.formula <- as.formula(
      paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))

# get the outcome vector 
y <- subsample$y  
# get the treatment assignment vector 
treat <- subsample$treat  
# get the design matrix (with treatment column)
X.tilde <- model.matrix(reg.formula, data = subsample)
# get the design matrix (no treatment column)
X <- model.matrix(treat.reg.formula, data = subsample)

result <- est_aipw(this.y=y, this.X.tilde=X.tilde, this.X=X, this.treat=treat, estimand=estimand, 
                   treat.reg.formula=treat.reg.formula, reg.formula=reg.formula)
```

The results is a list of estimation outputs including the ATE estimate (`result$est`), the corresponding variance (`result$V`) and the influence functions (`result$influence.function`). 

## federation using AIPW

### stable models 

In this example, we assume stable models across data sets. Our target estimand is ATE (`estimand="ATE"`). 

```{r}
# specify the working covariates 
# for now we assume stable models, so same set of covariates across data sets 
covariates <- c("X1", "X2", "X3")   

# initialize inputs for federated AIPW function
full_y <- vector(mode = "list", length = D)
full_X.tilde <- full_X <- vector(mode = "list", length = D)
full_treat <- vector(mode = "list", length = D)
full_reg.formula <- full_treat.reg.formula <- vector(mode = "list", length = D)

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  
  reg.formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))
  treat.reg.formula <- as.formula(
    paste("treat ~ 0 + intercept + ", paste(covariates, collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_treat[[ix]] <- subsample$treat
  full_X[[ix]] <- model.matrix(treat.reg.formula, data = subsample)
  full_X.tilde[[ix]] <- model.matrix(reg.formula, data = subsample)
  full_reg.formula[[ix]] <- reg.formula
  full_treat.reg.formula[[ix]] <- treat.reg.formula
}

# restricted federated AIPW estimation 
result <- poolAIPW(full_y=full_y, full_X.tilde=full_X.tilde, full_X=full_X, full_treat=full_treat, 
                   unstable=FALSE, 
                   full_reg.formula=full_reg.formula, full_treat.reg.formula=full_treat.reg.formula, 
                   estimand="ATE")
```

The returned summary table reports the pooled estimates, standard errors, t statistics and corresponding p-values of the ATE estimation.

```{r}
result
```

### unstable models 

In this example, we assume unstable models across data sets. Assume that we still use ATE as our target estimand (`estimand="ATE"`). 

```{r}
# specify the working covariates 
# we assume unstable models
covariates <- c("X1", "X2", "X3")   

# initialize inputs for federated AIPW function
full_y <- vector(mode = "list", length = D)
full_X.tilde <- full_X <- vector(mode = "list", length = D)
full_treat <- vector(mode = "list", length = D)
full_reg.formula <- full_treat.reg.formula <- vector(mode = "list", length = D)

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  
  reg.formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates, collapse = " + ")))
  treat.reg.formula <- as.formula(
    paste("treat ~ 0 + intercept + ", paste(covariates, collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_treat[[ix]] <- subsample$treat
  full_X[[ix]] <- model.matrix(treat.reg.formula, data = subsample)
  full_X.tilde[[ix]] <- model.matrix(reg.formula, data = subsample)
  full_reg.formula[[ix]] <- reg.formula
  full_treat.reg.formula[[ix]] <- treat.reg.formula
}

# unrestricted federated AIPW estimation 
result <- poolAIPW(full_y=full_y, full_X.tilde=full_X.tilde, full_X=full_X, full_treat=full_treat, 
                   unstable=TRUE, 
                   full_reg.formula=full_reg.formula, full_treat.reg.formula=full_treat.reg.formula, 
                   estimand="ATE")
```

The returned summary table reports the pooled estimates, standard errors, t statistics and corresponding p-values of the ATE estimation.

```{r}
result
```

We can specify different working models across data sets and then use unrestricted federated AIPW estimator: 

```{r}
# specify the working covariates 
# we assume unstable models which include different sets of covariates 
covariates <- list(c("X1", "X2", "X3"), c("X1", "X3"))

# initialize inputs for federated AIPW function
full_y <- vector(mode = "list", length = D)
full_X.tilde <- full_X <- vector(mode = "list", length = D)
full_treat <- vector(mode = "list", length = D)
full_reg.formula <- full_treat.reg.formula <- vector(mode = "list", length = D)

for (ix in 1:D) {
  subsample <- subsample_lst[[ix]]
  
  reg.formula <- as.formula(
    paste("y ~ 0 + intercept + treat + ", paste(covariates[[ix]], collapse = " + ")))
  treat.reg.formula <- as.formula(
    paste("treat ~ 0 + intercept + ", paste(covariates[[ix]], collapse = " + ")))
  
  full_y[[ix]] <- subsample$y
  full_treat[[ix]] <- subsample$treat
  full_X[[ix]] <- model.matrix(treat.reg.formula, data = subsample)
  full_X.tilde[[ix]] <- model.matrix(reg.formula, data = subsample)
  full_reg.formula[[ix]] <- reg.formula
  full_treat.reg.formula[[ix]] <- treat.reg.formula
}

# unrestricted federated AIPW estimation 
result <- poolAIPW(full_y=full_y, full_X.tilde=full_X.tilde, full_X=full_X, full_treat=full_treat, 
                   unstable=TRUE, 
                   full_reg.formula=full_reg.formula, full_treat.reg.formula=full_treat.reg.formula, 
                   estimand="ATE")
```

And here is the returned summary table of the federated ATE estimation:

```{r}
result
```